raptr: Representative and Adequate Prioritization Toolkit in R

### 2016-11-28 code/rmarkdown/article.Rmd
+ add in document statistics for article + reduce length of abstract, main text and figures to within limits

---
title: "raptr: Representative and Adequate Prioritization Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^{1,3}$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | $^3$The Nature Conservancy, South Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimization", "protected area"]
abstract: "

1. An underlying aim in conservation is to maximize the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, the long-term persistence of a species may depend on its ability to adapt to new climatic conditions. By conserving individuals with the ability to persist in a range of conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Despite this, current reserve selection methods overwhelmingly focus on securing an adequate proportion of features' geographic ranges and little attention has been directed towards capturing a representative sample.

2. To address this issue, we developed the \\texttt{raptr R} package to guide reserve selection using targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set "amount targets"--similar to conventional methods--to secure a target proportion of the features' geographic distributions. Additionally, users set "space targets" to secure a representative sample of variation in ecologically or evolutionary relevant attributes (eg. climatic or genetic variation). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' geographic distributions using amount targets--representing conventional methods--and compare them with solutions generated using amount and space targets.

@@

4. The \\texttt{raptr R} package provides a toolkit for making spatial conservation prioritizations that secure an adequate and representative sample of features. By targeting representativeness, prioritizations may have a greater chance of achieving long-term biodiversity persistence.

"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble.tex
fontsize: 11pt
documentclass: article
subparagraph: true
bibliography: references.bib
csl: reference-style.csl
---

Short running title: Representative and adequate prioritizations  
Abstract word count: 350 / 350  
Total word count: 6977 / 7000  
Number of references: 42  
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 4

>>>>

---
title: "raptr: Representative and Adequate Prioritization Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^{1,3}$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | $^3$The Nature Conservancy, South Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimization", "protected area"]
abstract: "
1. An underlying aim in conservation is to maximize the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, the long-term persistence of a species may depend on its ability to adapt to new climatic conditions. By conserving individuals with the ability to persist in a range of conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Despite this, current reserve selection methods overwhelmingly focus on securing an adequate proportion of features' geographic ranges and little attention has been directed towards capturing a representative sample.

2. To address this issue, we developed the \\texttt{raptr R} package to guide reserve selection using targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional methods--to secure a sufficient proportion of the features' geographic distributions. Additionally, users set \"space targets\" to secure a representative sample of variation in ecologically or evolutionary relevant attributes (eg. climatic or genetic variation). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions using just amount targets--representing conventional methods--and compare them with solutions generated using amount and space targets.

@@

4. The \\texttt{raptr R} package provides a toolkit for making spatial conservation prioritizations that secure an adequate and representative sample of features. By targeting representativeness, prioritizations may have a greater chance of achieving long-term biodiversity persistence.
"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble.tex
fontsize: 11pt
documentclass: article
subparagraph: true
bibliography: references.bib
csl: reference-style.csl
---

Short running title: Representative and adequate prioritizations  
Abstract word count: 345 / 350  
Total word count: 6999 / 7000  
Number of references: 42  
Number of figures: 8  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: 4

### 2016-11-28 code/rmarkdown/figures.Rmd
+ add in document statistics for article + reduce length of abstract, main text and figures to within limits

```{r, echo=FALSE, fig.height=2.35, fig.width=column.width, fig.cap='Distribution of the species used in the first case-study. Each panel shows a different species. Squares denote planning units. Planning units occupied by a species are shown in bright blue.'}
# prepare data
cs1.pus.occ.FPLY <- ldply(
	seq_along(cs1.params.LST[[MODE]]$common.names),
	function(i) {
		curr.DF <- cs1.pus.FPLY
		curr.DF$species <- cs1.params.LST[[MODE]]$common.names[i]
		curr.DF$Occupancy <- 'absent'
		curr.DF$Occupancy[which(curr.DF$id %in% filter(cs1.ru@data@pu.species.probabilities, species==i)$pu)] <- 'present'
		return(curr.DF)
	}
)
cs1.pus.occ.FPLY$species <- factor(as.character(cs1.pus.occ.FPLY$species), levels=cs1.params.LST[[MODE]]$common.names)
# make letter data.frame
cs1.pus.occ.letters.DF <- data.frame(
	x=unname(cs1.pus@bbox[1,1] + (diff(cs1.pus@bbox[1,])*0.06)),
	y=unname(cs1.pus@bbox[2,1] + (diff(cs1.pus@bbox[2,])*0.90)),
	species=unique(cs1.pus.occ.FPLY$species),
	letter=paste0('(', letters[seq_along(unique(cs1.pus.occ.FPLY$species))], ')')
)
# make plots
ggplot() +
	geom_polygon(data=aus.FPLY, aes(x=long, y=lat, group=group), fill='grey85', color='grey70') +
	geom_polygon(data=cs1.pus.occ.FPLY, aes(x=long, y=lat, group=group, fill=Occupancy), color='black', size=0.05) +
	geom_label(data=cs1.pus.occ.letters.DF, aes(label=letter, x=x, y=y), color='black', fill=alpha('white', 0.8), size=3) +
	theme_classic() +
	theme(axis.ticks=element_blank(), axis.text=element_blank(), axis.line=element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"),
		strip.background = element_rect(fill='grey20'),
		strip.text = element_text(color='white', size=8),
		axis.title=element_blank(), legend.position='none'
	) +
	scale_fill_manual(name='Occupancy', values=c('absent'='#132B43', 'present'='#56B1F7')) +
	coord_cartesian(xlim=cs1.pus@bbox[1,], ylim=cs1.pus@bbox[2,]) +
	facet_wrap(~ species, ncol=2)
```

```{r, echo=FALSE, fig.height=1.175, fig.width=column.width, fig.cap=paste0('Prioritizations for the first case study. Squares represent planning units. Dark green planning units were selected for protection. Panel (a) shows the planning units selected when using ',cs1.params.LST[[MODE]]$amount.target*100,' % amount targets. Panel (b) shows the planning units selected when using ',cs1.params.LST[[MODE]]$amount.target*100,' % amount targets and ',cs1.params.LST[[MODE]]$space.target*100,' % niche targets')}
# prepare data
cs1.pus.sel.FPLY <- ldply(
	names(cs1.prioritisations),
	function(x) {
		curr.DF <- cs1.pus.FPLY
		curr.DF$Prioritisation <- x
		curr.DF$Solution <- 'discarded'
		curr.DF$Solution[curr.DF$id %in% which(as.logical(selections(cs1.prioritisations[[x]])))] <- 'selected'
		return(curr.DF)
	}
)
cs1.pus.sel.FPLY$Prioritisation <- factor(as.character(cs1.pus.sel.FPLY$Prioritisation), levels=names(cs1.prioritisations))
# make letter data.frame
cs1.pus.sel.letters.DF <- data.frame(
	x=unname(cs1.pus@bbox[1,1] + (diff(cs1.pus@bbox[1,])*0.06)),
	y=unname(cs1.pus@bbox[2,1] + (diff(cs1.pus@bbox[2,])*0.90)),
	Prioritisation=names(cs1.prioritisations),
	letter=paste0('(', letters[1:2], ')')
)
# make plots
ggplot() +
	geom_polygon(data=aus.FPLY, aes(x=long, y=lat, group=group), fill='grey85', color='grey70') +
	geom_polygon(data=cs1.pus.sel.FPLY, aes(x=long, y=lat, group=group, fill=Solution), color='black', size=0.1) +
	geom_label(data=cs1.pus.sel.letters.DF, aes(label=letter, x=x, y=y), color='black', fill=alpha('white', 0.8), size=3) +
	theme_classic() +
	theme(axis.ticks=element_blank(), axis.text=element_blank(), axis.line=element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"),
		strip.background = element_rect(fill='grey20'),
		strip.text = element_text(color='white', size=8),
		axis.title=element_blank(), legend.position='none'
	) +
	scale_fill_manual(name='Solution', values=c('discarded'='#f7fcfd', 'selected'='#00441b')) +
	coord_cartesian(xlim=cs1.pus@bbox[1,], ylim=cs1.pus@bbox[2,]) +
	facet_wrap(~ Prioritisation, ncol=2)
```

```{r, echo=FALSE, fig.height=2.5, fig.width=column.width, fig.cap=paste0('Data used for the second case-study. Squares denote planning units. Panel (a) shows all grid cells surveyed by the IntraBioDiv project. Grid cells occupied by ',tolower(cs2.params.LST[[MODE]]$common.name),' are shown in bright blue. The subsequent panels contain only the grid cells that were occupied by the species. Panel (b) shows the opportunity cost of each planning unit (estimated as the total human population density). Panels (c--d) show the spatial distribution of the ordinated genetic data. These values describe the typical genetic characteristics of individuals in each planning unit. Planning units with similar values/colors contain individuals with similar loci polymorphisms.')}
# prepare full grid data
cs2.spp.grid.FPLY <- cs2.grid.FPLY
cs2.spp.grid.FPLY$Title <- 'Occupancy'
cs2.spp.grid.FPLY$Occupancy <- 'absent'
cs2.spp.grid.FPLY$Occupancy[which(cs2.grid.PLY@data[[cs2.params.LST[[MODE]]$species.name]][as.numeric(cs2.spp.grid.FPLY$id)]==1)] <- 'present'

@@

```{r, echo=FALSE, fig.height=2.5, fig.width=column.width, fig.cap=paste0('Prioritizations for the second case study. Panels (a--b) show prioritisations generated using different parameters. Squares denote planning units shown in geographic space. Dark green planning units were selected for protection. Panel (a) shows the planning units selected when using ',cs2.params.LST[[MODE]]$amount.target*100,' % amount targets. Panel (b) shows the planning units selected when using ',cs2.params.LST[[MODE]]$amount.target*100,' % amount targets and ',cs2.params.LST[[MODE]]$genetic.target*100,' % genetic targets. Panels (c--d) show the solutions in the genetic space. Each point corresponds to a planning unit. The coordinates of the points represent the typical genetic characeristics of individuals sampled in that planning unit (based on an NMDS of the binary loci data). Planning units associated with points that are closer togeather contain individuals with more similar genetic characteristics than planning units that are further apart.')}
# prepare spatial data
cs2.sel.grid.sub.FPLY <- cs2.grid.sub.FPLY
cs2.sel.grid.sub.FPLY$Title1 <-'Amount target'
cs2.sel.grid.sub.FPLY$Title2 <- 'Amount & genetic target'

>>>>

```{r, echo=FALSE, fig.height=2.35, fig.width=column.width, fig.cap='Distribution of the species used in the first case study. Each panel shows a different species. Squares denote planning units. Planning units occupied by a species are shown in bright blue.'}
# prepare data
cs1.pus.occ.FPLY <- ldply(
	seq_along(cs1.params.LST[[MODE]]$common.names),
	function(i) {
		curr.DF <- cs1.pus.FPLY
		curr.DF$species <- cs1.params.LST[[MODE]]$common.names[i]
		curr.DF$Occupancy <- 'absent'
		curr.DF$Occupancy[which(curr.DF$id %in% filter(cs1.ru@data@pu.species.probabilities, species==i)$pu)] <- 'present'
		return(curr.DF)
	}
)
cs1.pus.occ.FPLY$species <- factor(as.character(cs1.pus.occ.FPLY$species), levels=cs1.params.LST[[MODE]]$common.names)
# make letter data.frame
cs1.pus.occ.letters.DF <- data.frame(
	x=unname(cs1.pus@bbox[1,1] + (diff(cs1.pus@bbox[1,])*0.06)),
	y=unname(cs1.pus@bbox[2,1] + (diff(cs1.pus@bbox[2,])*0.90)),
	species=unique(cs1.pus.occ.FPLY$species),
	letter=paste0('(', letters[seq_along(unique(cs1.pus.occ.FPLY$species))], ')')
)
# make plots
ggplot() +
	geom_polygon(data=aus.FPLY, aes(x=long, y=lat, group=group), fill='grey85', color='grey70') +
	geom_polygon(data=cs1.pus.occ.FPLY, aes(x=long, y=lat, group=group, fill=Occupancy), color='black', size=0.05) +
	geom_label(data=cs1.pus.occ.letters.DF, aes(label=letter, x=x, y=y), color='black', fill=alpha('white', 0.8), size=3) +
	theme_classic() +
	theme(axis.ticks=element_blank(), axis.text=element_blank(), axis.line=element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"),
		strip.background = element_rect(fill='grey20'),
		strip.text = element_text(color='white', size=8),
		axis.title=element_blank(), legend.position='none'
	) +
	scale_fill_manual(name='Occupancy', values=c('absent'='#132B43', 'present'='#56B1F7')) +
	coord_cartesian(xlim=cs1.pus@bbox[1,], ylim=cs1.pus@bbox[2,]) +
	facet_wrap(~ species, ncol=2)
```

```{r, echo=FALSE, fig.height=1.175, fig.width=column.width, fig.cap=paste0('Prioritizations for the first case study. Squares represent planning units. Dark green planning units were selected for protection. Panel (a) shows the planning units selected when using ',cs1.params.LST[[MODE]]$amount.target*100,' % amount targets. Panel (b) shows the planning units selected when using ',cs1.params.LST[[MODE]]$amount.target*100,' % amount targets and ',cs1.params.LST[[MODE]]$space.target*100,' % niche targets.')}
# prepare data
cs1.pus.sel.FPLY <- ldply(
	names(cs1.prioritisations),
	function(x) {
		curr.DF <- cs1.pus.FPLY
		curr.DF$Prioritisation <- x
		curr.DF$Solution <- 'discarded'
		curr.DF$Solution[curr.DF$id %in% which(as.logical(selections(cs1.prioritisations[[x]])))] <- 'selected'
		return(curr.DF)
	}
)
cs1.pus.sel.FPLY$Prioritisation <- factor(as.character(cs1.pus.sel.FPLY$Prioritisation), levels=names(cs1.prioritisations))
# make letter data.frame
cs1.pus.sel.letters.DF <- data.frame(
	x=unname(cs1.pus@bbox[1,1] + (diff(cs1.pus@bbox[1,])*0.06)),
	y=unname(cs1.pus@bbox[2,1] + (diff(cs1.pus@bbox[2,])*0.90)),
	Prioritisation=names(cs1.prioritisations),
	letter=paste0('(', letters[1:2], ')')
)
# make plots
ggplot() +
	geom_polygon(data=aus.FPLY, aes(x=long, y=lat, group=group), fill='grey85', color='grey70') +
	geom_polygon(data=cs1.pus.sel.FPLY, aes(x=long, y=lat, group=group, fill=Solution), color='black', size=0.1) +
	geom_label(data=cs1.pus.sel.letters.DF, aes(label=letter, x=x, y=y), color='black', fill=alpha('white', 0.8), size=3) +
	theme_classic() +
	theme(axis.ticks=element_blank(), axis.text=element_blank(), axis.line=element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"),
		strip.background = element_rect(fill='grey20'),
		strip.text = element_text(color='white', size=8),
		axis.title=element_blank(), legend.position='none'
	) +
	scale_fill_manual(name='Solution', values=c('discarded'='#f7fcfd', 'selected'='#00441b')) +
	coord_cartesian(xlim=cs1.pus@bbox[1,], ylim=cs1.pus@bbox[2,]) +
	facet_wrap(~ Prioritisation, ncol=2)
```

```{r, echo=FALSE, fig.height=2.5, fig.width=column.width, fig.cap=paste0('Data used for the second case study. Squares denote planning units. Panel (a) shows all grid cells surveyed by the IntraBioDiv project. Grid cells occupied by ',tolower(cs2.params.LST[[MODE]]$common.name),' are shown in bright blue. The subsequent panels contain only show occupied grid cells. Panel (b) shows the opportunity cost of each planning unit (estimated as the total human population density). Panels (c--d) show the spatial distribution of the ordinated genetic data. These values describe the typical genetic characteristics of individuals in each planning unit. Planning units with similar values/colors contain individuals with similar loci polymorphisms.')}
# prepare full grid data
cs2.spp.grid.FPLY <- cs2.grid.FPLY
cs2.spp.grid.FPLY$Title <- 'Occupancy'
cs2.spp.grid.FPLY$Occupancy <- 'absent'
cs2.spp.grid.FPLY$Occupancy[which(cs2.grid.PLY@data[[cs2.params.LST[[MODE]]$species.name]][as.numeric(cs2.spp.grid.FPLY$id)]==1)] <- 'present'

@@

```{r, echo=FALSE, fig.height=2.5, fig.width=column.width, fig.cap=paste0('Prioritizations for the second case study. Panels (a--b) show prioritizations generated using different parameters. Squares denote planning units shown in geographic space. Dark green planning units were selected for protection. Panel (a) shows the planning units selected when using ',cs2.params.LST[[MODE]]$amount.target*100,' % amount targets. Panel (b) shows the planning units selected when using ',cs2.params.LST[[MODE]]$amount.target*100,' % amount targets and ',cs2.params.LST[[MODE]]$genetic.target*100,' % genetic targets. Panels (c--d) show the solutions in the genetic space. Each point corresponds to a planning unit. The coordinates of the points represent the typical genetic characteristics of individuals sampled in that planning unit (based on an NMDS of the binary loci data). Planning units associated with points that are closer together contain individuals with more similar genetic characteristics than planning units that are further apart.')}
# prepare spatial data
cs2.sel.grid.sub.FPLY <- cs2.grid.sub.FPLY
cs2.sel.grid.sub.FPLY$Title1 <-'Amount target'
cs2.sel.grid.sub.FPLY$Title2 <- 'Amount & genetic target'

### 2016-11-28 code/rmarkdown/text.Rmd
+ add in document statistics for article + reduce length of abstract, main text and figures to within limits

## Optimization
The unreliable formulation is non-linear. However, the non-linear components can be linearized using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearized using methods described by Beyer et al. [-@r426]. Linearized versions of the problems can be solved using commercial exact algorithms. The \texttt{raptr} R package provides functions to express conservation planning data as optimization problems using linearized versions of the unreliable and reliable formulations (see Appendix S1 for formulation). These optimization problems can then be solved to generate prioritizations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

To understand the behavior of the unreliable problem and showcase its value, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritizations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can gauge the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This specific configuration is an artifact of the \texttt{Gurobi} solver used to solve the problem. In fact, for this species, all prioritizations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply siting a planning unit in the centroid of the species' demand points.

>>>>

## Optimization
The unreliable formulation is non-linear. However, the non-linear components can be linearized using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearized using methods described by Beyer et al. [-@r426]. The \texttt{raptr} R package provides functions to express conservation planning data as optimization problems using linearized versions of the unreliable and reliable formulations (see Appendix S1 for formulation), and solve them using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

@@

To understand and showcase the behavior of the unreliable problem, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritizations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can gauge the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

@@

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This specific configuration is an artifact of the method used to solve the problem. In fact, for this species, all prioritizations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of additional criteria, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply siting a planning unit in the centroid of the species' demand points.
